\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{booktabs}
\usepackage{microtype}
\usepackage{natbib}
\usepackage[hidelinks]{hyperref}

\newcommand{\R}{\mathbb{R}}
\newcommand{\Sig}{\Sigma}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\Aset}{\mathcal{A}}
\newcommand{\Sset}{\mathcal{S}}
\newcommand{\trans}{^{\top}}
\newcommand{\lb}{\ell}
\newcommand{\ub}{u}

\newtheorem{proposition}{Proposition}
\newtheorem{definition}{Definition}

\title{Following the Solution:\\
Parametric Active-Set Homotopies in Statistics, Optimization, and Finance}

\author{[Authors]\thanks{Perspective draft for \emph{Statistical Science}.}}
\date{Draft, \today}

\begin{document}
\maketitle

\begin{abstract}
A striking number of methods across statistics, optimization, and finance do not
return a single solution but trace an entire \emph{path} of solutions as one scalar
parameter is swept: least angle regression and the LASSO follow the
$\ell_1$-regularised coefficients along the penalty; the support-vector-machine
regularization path follows the dual variables along the cost; explicit model
predictive control precomputes the control law as the state varies; and Markowitz's
Critical Line Algorithm, sixty years older than any of them, follows the optimal
portfolio along risk aversion. Each community arrived at the same device
independently and named it differently. We argue that they are one device. When a
strictly convex quadratic is followed along a single parameter under polyhedral
structure, the solution map is piecewise linear and is traced exactly by an
active-set walk from one breakpoint to the next; everything else is interface. We
develop this common scheme, place the major instances inside it, and use the
cleanest pair, the LASSO and the Critical Line Algorithm, to show that the identity
is exact rather than analogical, surviving general linear equality and inequality
constraints. We then separate what is genuinely shared (the homotopy, the
simplex-inherited anti-cycling, the access to data through a quadratic form) from
what is particular to each field (whether the active set is imposed by explicit
constraints or induced by a penalty, whether variables only enter or also leave, and
what structure the quadratic form carries). A short discussion collects the open
questions the unified view makes precise, chiefly the gap between the linear path
length seen in practice and the exponential one possible in theory.
\end{abstract}

\section{A device with many names}

Modern statistical computing is full of \emph{paths}. We rarely want the LASSO fit
at one penalty; we want the whole trajectory of coefficients as the penalty relaxes,
because the trajectory is where model selection, stability, and interpretation live.
The same is true far outside statistics. A portfolio manager wants not one portfolio
but the efficient frontier, the trajectory of optimal allocations as the appetite for
return varies. A control engineer wants not one control action but the law that maps
every state to its optimal action. In each case the object of interest is a curve in
solution space indexed by a scalar.

These curves were discovered, and rediscovered, in isolation. \citet{markowitz1956,
markowitz1959} gave the Critical Line Algorithm (CLA) for the efficient frontier in
the 1950s. \citet{osborne2000} and then \citet{efron2004} gave the homotopy and
least angle regression (LARS) for the LASSO of \citet{tibshirani1996} around 2000.
\citet{hastie2004svm} traced the entire regularization path of the support vector
machine; \citet{zou2005} the elastic net; \citet{tibshirani2011} the generalized
lasso, including the fused lasso and trend filtering. In control,
\citet{bemporad2002} and \citet{tondel2003} precompute the constrained
linear-quadratic regulator as an explicit piecewise-affine law by multi-parametric
quadratic programming. The literatures cite within themselves and seldom across.

Our thesis is that these are not analogies but instances of one scheme, and that
saying so precisely is worthwhile on three counts. It is unifying: a reader fluent in
one field can read the others at sight. It is generative: machinery developed for one
instance transfers, and we will see an exact constrained LASSO path fall out of the
portfolio construction. And it is clarifying: stating the common core exposes what is
genuinely particular to each problem, which is usually less than its native
presentation suggests. The clearest evidence is a pair that look maximally different,
a portfolio optimizer and a sparse regression, that turn out to run the identical
algorithm. We build to that identity and then step back from it.

\section{The common scheme}
\label{sec:scheme}

Fix a strictly convex quadratic objective whose linear part carries a scalar
parameter $\lambda$, minimised over a polyhedron:
\begin{equation}
\label{eq:pqp}
x(\lambda) \;=\; \arg\min_{x \in \R^n}\ \tfrac12\, x\trans H x - \lambda\, d\trans x
\quad\text{subject to}\quad A x = b,\ \ G x \le h ,
\end{equation}
with $H \succ 0$. (A polyhedral \emph{penalty}, such as $\lambda\|x\|_1$, fits the
same template through its subgradient, as we make explicit below.) This is a
\emph{one-parameter parametric quadratic program}. Its solution map has a structure
that is the entire story.

\paragraph{Piecewise linearity.}
Partition the constraints at a given $\lambda$ into \emph{active} (holding with
equality) and \emph{inactive}. On the set of $\lambda$ for which the active set is
fixed, the Karush--Kuhn--Tucker (KKT) conditions are a linear system whose matrix is
constant (it depends only on $H$ and the active constraint normals) and whose
right-hand side is affine in $\lambda$ (only the term $\lambda d$ moves). Hence on
each such region the optimiser is affine,
\begin{equation}
\label{eq:affine}
x(\lambda) \;=\; \alpha + \lambda\,\beta ,
\end{equation}
and the regions partition the $\lambda$-axis into intervals separated by
\emph{breakpoints}, the values at which the active set must change. \citet{rosset2007}
gave the clean sufficient condition: a quadratic (or piecewise-quadratic) loss
together with a polyhedral, i.e.\ piecewise-linear, penalty or constraint forces a
piecewise-linear path. In the optimization and control literature the same fact is
the statement that the solution of \eqref{eq:pqp} is piecewise affine over a
polyhedral partition of the parameter space into \emph{critical regions}
\citep{bemporad2002}.

\paragraph{The homotopy.}
Because the solution is affine on each region and continuous across breakpoints, the
entire path is recovered by a homotopy: start at a vertex where the active set is
known, follow the affine segment \eqref{eq:affine} until the first event (an inactive
constraint becoming active, or an active one releasing), update the active set by
that single event, and repeat. Each step is one linear solve against the active
block of $H$, bordered by the active constraint normals. This is exactly the
turning-point loop of the CLA, the LARS recursion, and the critical-region traversal
of explicit MPC, written once.

\paragraph{Anti-cycling.}
The homotopy is a continuation method, but its event selection, ``the smallest
positive step to the next breakpoint,'' is the simplex method's ratio test in
disguise, and it inherits the simplex method's degeneracy hazard: at a point where
several events tie, a naive rule can cycle. The remedy is also inherited, a
lowest-index (Bland) tie-break that makes the walk deterministic. That the same
anti-cycling device appears in LARS tie handling and in CLA implementations is not a
coincidence; it is the shared simplex ancestry showing through.

We take \eqref{eq:pqp}, \eqref{eq:affine}, and the homotopy as the \emph{common
scheme}. The remainder of the paper is what fills in $H$, $d$, and the constraints.

\section{A tour of the instances}
\label{sec:tour}

\paragraph{LARS and the LASSO.}
For a response $y \in \R^m$ and design $X \in \R^{m\times n}$, the LASSO
\citep{tibshirani1996} solves $\min_\beta \tfrac12\|y - X\beta\|_2^2 +
\lambda\|\beta\|_1$. Here $H = X\trans X$ and $d = X\trans y$. The $\ell_1$ penalty is
polyhedral; its subgradient pins each inactive coordinate inside $[-\lambda,\lambda]$
in correlation and fixes the sign of each active one. On a segment with support
$\Aset$ and signs $s_\Aset$ the stationarity condition reads $X_\Aset\trans X_\Aset
\beta_\Aset = X_\Aset\trans y - \lambda s_\Aset$, affine in $\lambda$, and the path is
piecewise linear \citep{osborne2000, efron2004}. The active set changes when an
inactive coordinate's correlation reaches $\lambda$ (enter) or an active coefficient
reaches zero (leave).

\paragraph{The regularization-path family.}
The same template generated a decade of statistics. The support vector machine's
regularization path \citep{hastie2004svm} follows the dual variables as the cost
parameter varies, again piecewise linear, again traced by an active-set homotopy. The
elastic net \citep{zou2005} adds a ridge term, which changes $H$ to $X\trans X +
\eta I$ but not the scheme. The generalized lasso \citep{tibshirani2011}, with
penalty $\lambda\|D\beta\|_1$ for a structured $D$ (the fused lasso, trend filtering,
and graph penalties as special cases), has a piecewise-linear \emph{dual} path traced
the same way. In each case a different penalty or loss instantiates
\eqref{eq:pqp}, and the homotopy is reused.

\paragraph{The Critical Line Algorithm.}
For weights $w \in \R^n$, expected returns $\mu$, and covariance $\Sig \succ 0$, the
CLA traces
\begin{equation}
\label{eq:cla}
\min_{w}\ \tfrac12\, w\trans \Sig\, w - \lambda\, \mu\trans w
\quad\text{s.t.}\quad \mathbf{1}\trans w = 1,\ \ \lb \le w \le \ub ,
\end{equation}
as $\lambda$ runs from $+\infty$ (maximum return) to $0$ (minimum variance). Here
$H = \Sig$ and $d = \mu$; the active set is the set of assets pinned at a box bound;
the events are a free weight reaching a bound or a bound multiplier releasing. The
locus of optimisers is the efficient frontier. This is the oldest instance of the
scheme by four decades, and the one statisticians rarely meet.

\paragraph{Explicit model predictive control.}
The constrained linear-quadratic regulator, solved offline as a function of the
state, is a multi-parametric QP whose solution is a piecewise-affine control law over
critical regions \citep{bemporad2002, tondel2003}. The parameter there is a vector
(the state), so the partition is of a polyhedron rather than an interval, but the
one-parameter case is exactly \eqref{eq:pqp}, and the region-traversal algorithms are
the multidimensional cousins of the homotopy.

\section{The sharpest identity: the CLA is a constrained LASSO}
\label{sec:identity}

Among these instances, two look maximally unalike and are in fact the same
computation: the LASSO and the CLA. Comparing their segment equations, the covariance
$\Sig$ plays the role of the Gram matrix $X\trans X$, and the expected-return vector
$\mu$ the role of the correlations $X\trans y$. Under that single substitution the two
homotopies coincide term for term (Table~\ref{tab:dict}).

\begin{table}[t]
\centering
\begin{tabular}{lll}
\toprule
Ingredient & Critical Line Algorithm & LARS / LASSO homotopy \\
\midrule
Swept parameter & risk aversion $\lambda:\infty\!\to\!0$ & penalty $\lambda:\infty\!\to\!0$ \\
Quadratic form $H$ & covariance $\Sig$ & Gram matrix $X\trans X$ \\
Linear term $d$ & expected returns $\mu$ & correlations $X\trans y$ \\
Solution path & $w(\lambda)=\alpha+\lambda\beta$ & $\beta(\lambda)=\alpha-\lambda\beta$ \\
Active set & free assets (off the bounds) & support (nonzero $\beta_j$) \\
Enter event & blocked multiplier changes sign & correlation reaches $\lambda$ \\
Leave event & free weight reaches a box bound & coefficient reaches zero \\
Step length & \multicolumn{2}{l}{smallest positive step to the next breakpoint} \\
Degeneracy & lowest-index tie-break & lowest-index tie-break \\
\bottomrule
\end{tabular}
\caption{The CLA and the LARS/LASSO homotopy under the substitution
$(\Sig,\mu)\leftrightarrow(X\trans X,\,X\trans y)$.}
\label{tab:dict}
\end{table}

\paragraph{The identity survives general constraints.}
The correspondence above is the textbook (budget-only or unconstrained) one. Its real
content is that it extends, identically on both sides, to general linear constraints.
Add an equality system $Ax=b$ and inequalities $Gx \le h$ to either problem, as in
\eqref{eq:pqp}. An \emph{active} inequality row, held at $g_i\trans x = h_i$, enters
the reduced KKT system exactly as an equality row does: stacking $C = [\,A;\,
G_{\Sset}\,]$ over the active rows recovers the equality-only solve with $A$ replaced
by $C$, so each step remains one solve against the active block of $H$ bordered by an
$(m + |\Sset|)$-dimensional Schur complement. Box bounds and coefficient signs stay a
separate, cheaper per-variable active set. The consequence for statistics is concrete:
\eqref{eq:pqp} traces, exactly and in one pass, the LASSO path under arbitrary
$G\beta \le h$, under non-negativity $\beta \ge 0$, and under linear equalities,
constrained paths that coordinate descent and proximal methods only approximate on a
grid and that LARS does not provide. The dictionary extends accordingly: a long-only
portfolio is a non-negative LASSO; the budget $\mathbf{1}\trans w = 1$ is a linear
equality on $\beta$; a factor-neutrality row is a linear equality; a group-exposure
cap is an inequality row.

\paragraph{Constraints versus penalties.}
The one structural difference is illuminating rather than cosmetic. In the CLA the
active set is carved by \emph{explicit} polyhedral constraints, and the events are
weights reaching bounds. In the LASSO it is \emph{induced} by the $\ell_1$ penalty
through its subgradient, and sparsity is emergent geometry rather than a constraint
the modeller writes. The same polyhedral mechanism runs in opposite directions, one
imposing the faces and the other discovering them. \citet{brodie2009} located the
hinge: adding a gross-exposure constraint $\|w\|_1 \le c$ to the Markowitz problem
makes it a LASSO, with $c$ promoting sparse, stable portfolios. For a long-only,
fully invested portfolio the constraint is silent, since the budget already fixes
$\|w\|_1 = 1$; admit short positions under a gross-exposure cap and the explicit
constraint and the induced penalty describe the same polyhedron, and the two paths are
not merely similar but identical.

\section{Shared scaffolding, particular features}
\label{sec:taxonomy}

Having seen the instances coincide, it is worth saying crisply what is shared and
what is not. The shared part is the whole of Section~\ref{sec:scheme}: the parametric
QP, the affine segment, the homotopy, and the simplex-inherited anti-cycling. The
particular features are few, and they organise the family.

\emph{Imposed versus induced active sets.} As above, constraints (CLA, constrained
regression) impose faces; penalties (LASSO, generalized lasso, SVM) induce them. The
gross-exposure case shows the two are interconvertible, but the computational
ergonomics differ, which is why each field built its own version.

\emph{Monotone versus non-monotone active sets.} Forward selection and forward LARS
only \emph{add} variables. The LASSO modification, and the CLA throughout, both add
and drop, because coefficients can return to zero and weights can leave and re-enter
the bounds as the parameter moves. The CLA therefore corresponds to the full LASSO
homotopy, not to forward LARS.

\emph{The structure of $H$.} The scheme touches the data only through the quadratic
form $H$. What $H$ \emph{is} determines both interpretation and cost: a covariance in
finance, a Gram matrix in regression, a kernel matrix in the SVM path. Crucially, the
homotopy never needs $H$ explicitly; it needs only a few linear-algebraic operations
on it, a point we return to in Section~\ref{sec:computation}, and structured $H$ (a
factor model, a kernel) is the lever for scale.

\emph{Degeneracy.} Ties and near-rank-deficiency are the recurring practical hazard,
and the inherited Bland rule is a partial answer: it guarantees determinism but, as
in the simplex method, the worst-case behavior on adversarial degenerate instances is
not fully tamed. This is shared across the family and is the locus of most of the open
questions in Section~\ref{sec:open}.

\section{Computation: one engine, many problems}
\label{sec:computation}

The unified view has a software corollary that is more than tidy. Because every step
of the homotopy accesses $H$ only through a matrix--vector product, a solve against
the active block, and an active-to-inactive cross product, a single implementation can
serve every instance: it is told the quadratic form (as an operator, not a matrix) and
what an event means, and it traces the path. The same loop that follows the efficient
frontier follows the LASSO path when handed $X\trans X$ and $X\trans y$; the operator
it calls is named for a generic quadratic form, not for the covariance, for exactly
this reason. The \texttt{cvxcla} package \citep{cvxcla} realises this, and we use it
for two deliberately light confirmations that the identity of
Section~\ref{sec:identity} is exact and not merely structural. On a standard design,
its traced path agrees with \texttt{scikit-learn}'s \texttt{lars\_path} to
$3.2\times10^{-15}$ in the coefficients over the same breakpoints; under a
non-negativity and an inequality constraint, its constrained path matches a reference
quadratic-program solver, solved independently at each breakpoint, to
$\approx 6\times10^{-10}$.

Treating $H$ as an operator is also where the scheme meets scale. A structured
quadratic form, a diagonal-plus-low-rank factor covariance in finance or a kernel in
learning, supplies the same few operations far more cheaply than a dense $H$ would,
without any change to the homotopy. The portfolio analogue of a structured or
kernelised design is a factor risk model, and it is the route by which the
sixty-year-old algorithm reaches problem sizes its original dense formulation could
not.

\section{What we still do not know}
\label{sec:open}

The unified view sharpens several questions that each field has met in its own
dialect.

\emph{How long is the path?} Empirically the number of breakpoints is linear in the
problem size for both the LASSO and the CLA, and this is what makes the homotopy
attractive against gridding. In the worst case it is not: the lasso path can have a
number of breakpoints exponential in the dimension \citep{mairal2012}, mirroring the
exponential critical-region counts possible in multi-parametric QP. Characterising the
gap between the benign average and the malignant worst case, and the data conditions
that separate them, is open and shared.

\emph{When is the homotopy the right tool?} For a single penalty value, coordinate
descent or a general QP solver is often faster than tracing to that value; the
homotopy wins when the whole path, or an exact breakpoint structure, is wanted. The
unified view makes the trade-off precise but does not resolve where the crossover lies
as a function of $H$'s structure and conditioning.

\emph{How far does exactness reach?} Piecewise linearity requires a quadratic (or
piecewise-quadratic) loss; logistic and other losses give only piecewise-smooth
paths, traced approximately by predictor--corrector continuation. How much of the
clean active-set story survives, and with what guarantees, is partially mapped.

\emph{Degeneracy at scale.} The Bland rule guarantees termination but not graceful
behavior on the most degenerate large instances, precisely the regime where structured
$H$ makes large problems reachable. A worst-case-robust tie-break, or a proof that the
benign behavior is generic, would matter to every instance at once.

\section{Conclusion}

Statistics, optimization, and finance each discovered that following the optimiser of
a parametrised convex quadratic produces a piecewise-linear path that an active set
suffices to trace, and each named the discovery after its own problem. Seen together,
least angle regression, the regularization-path family, the explicit linear-quadratic
regulator, and the Critical Line Algorithm are one homotopy wearing different clothes,
and the LASSO and the CLA are the same computation under a one-line substitution that
survives general constraints. What is particular to each, an imposed versus an induced
active set, a monotone versus a non-monotone one, a covariance versus a Gram or kernel
matrix, is a short list against a large shared core. Recognising the core lets a
result or an implementation built for one instance serve the rest, and turns four
parallel literatures on path-following into one.

\begin{thebibliography}{Markowitz(1959)}

\bibitem[Bemporad et~al.(2002)]{bemporad2002} Bemporad, A., Morari, M., Dua, V., and
Pistikopoulos, E.~N. (2002). The explicit linear quadratic regulator for constrained
systems. \emph{Automatica}, 38(1), 3--20.

\bibitem[Brodie et~al.(2009)]{brodie2009} Brodie, J., Daubechies, I., De Mol, C.,
Giannone, D., and Loris, I. (2009). Sparse and stable Markowitz portfolios.
\emph{Proceedings of the National Academy of Sciences}, 106(30), 12267--12272.

\bibitem[Authors(2025)]{cvxcla} Authors (2025). \texttt{cvxcla}: the Critical Line
Algorithm with a covariance-operator interface. Companion software paper.

\bibitem[Efron et~al.(2004)]{efron2004} Efron, B., Hastie, T., Johnstone, I., and
Tibshirani, R. (2004). Least angle regression. \emph{The Annals of Statistics},
32(2), 407--499.

\bibitem[Hastie et~al.(2004)]{hastie2004svm} Hastie, T., Rosset, S., Tibshirani, R.,
and Zhu, J. (2004). The entire regularization path for the support vector machine.
\emph{Journal of Machine Learning Research}, 5, 1391--1415.

\bibitem[Mairal and Yu(2012)]{mairal2012} Mairal, J. and Yu, B. (2012). Complexity
analysis of the lasso regularization path. \emph{Proceedings of the 29th
International Conference on Machine Learning (ICML)}.

\bibitem[Markowitz(1956)]{markowitz1956} Markowitz, H.~M. (1956). The optimization of
a quadratic function subject to linear constraints. \emph{Naval Research Logistics
Quarterly}, 3(1--2), 111--133.

\bibitem[Markowitz(1959)]{markowitz1959} Markowitz, H.~M. (1959). \emph{Portfolio
Selection: Efficient Diversification of Investments}. Wiley, New York.

\bibitem[Osborne et~al.(2000)]{osborne2000} Osborne, M.~R., Presnell, B., and
Turlach, B.~A. (2000). A new approach to variable selection in least squares
problems. \emph{IMA Journal of Numerical Analysis}, 20(3), 389--403.

\bibitem[Rosset and Zhu(2007)]{rosset2007} Rosset, S. and Zhu, J. (2007). Piecewise
linear regularized solution paths. \emph{The Annals of Statistics}, 35(3),
1012--1030.

\bibitem[Tibshirani(1996)]{tibshirani1996} Tibshirani, R. (1996). Regression
shrinkage and selection via the lasso. \emph{Journal of the Royal Statistical
Society, Series B}, 58(1), 267--288.

\bibitem[Tibshirani and Taylor(2011)]{tibshirani2011} Tibshirani, R.~J. and Taylor,
J. (2011). The solution path of the generalized lasso. \emph{The Annals of
Statistics}, 39(3), 1335--1371.

\bibitem[T{\o}ndel et~al.(2003)]{tondel2003} T{\o}ndel, P., Johansen, T.~A., and
Bemporad, A. (2003). An algorithm for multi-parametric quadratic programming and
explicit MPC solutions. \emph{Automatica}, 39(3), 489--497.

\bibitem[Zou and Hastie(2005)]{zou2005} Zou, H. and Hastie, T. (2005). Regularization
and variable selection via the elastic net. \emph{Journal of the Royal Statistical
Society, Series B}, 67(2), 301--320.

\end{thebibliography}

\end{document}
